---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit README.Rmd, not README.md. -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

pkgload::load_all(
  ".",
  quiet = TRUE
)
```

# phyloPatch

`phyloPatch` implements patch-aware ancestral-state reconstruction
for continuous traits on rooted phylogenetic trees containing one or
more specified branch-localized shifts.

The package is designed for situations in which a localized large
trait change occurs on a particular phylogenetic branch and the
downstream observations may influence ancestral-state reconstruction
elsewhere in the tree.

The central idea is to separate the shift-associated descendant region
from the remainder of the phylogeny, perform ancestral-state
reconstruction on the resulting components, reconstruct the boundary
node connecting those components, and finally merge all estimates back
onto the internal nodes of the original tree.

## Core workflow

For a specified shift edge from parent node \(I\) to child node \(J\),
`phyloPatch` performs the following operations:

1. validates the phylogeny, observed terminal states, shift edges, and
   ancestral-state reconstruction backend;
2. assigns stable node identifiers before pruning or subtree extraction;
3. identifies the descendant patch associated with each shift edge;
4. separates the phylogeny into a mother component and one or more patch
   components;
5. performs ancestral-state reconstruction independently on eligible
   components;
6. reconstructs the designated boundary nodes using branch lengths from
   the original phylogeny; and
7. merges component and boundary estimates back onto the complete set of
   internal nodes of the original tree.

The primary user interface is:

```r
patch_asr(
  tree,
  states,
  shifts,
  asr_fun
)
```

## Current scope

The current implementation supports:

- rooted phylogenetic trees;
- fully bifurcating topologies;
- finite and strictly positive branch lengths;
- continuous traits;
- one finite observed trait value for every terminal taxon;
- internal or terminal shift children;
- one or more compatible shift edges; and
- multiple shifts whose descendant patch sets are pairwise disjoint.

The shift parent must be an internal node and cannot be the root.

Nested or overlapping patch regions are not currently supported.

## Ancestral-state reconstruction backend

`phyloPatch` is not tied to one particular ancestral-state
reconstruction method.

Users provide an ancestral-state reconstruction function with the
interface:

```r
asr_fun(tree, states)
```

For each phylogenetic component:

- `tree` is the component phylogeny;
- `states` is a named numeric vector ordered exactly as
  `tree$tip.label`; and
- the function must return one finite ancestral-state estimate for every
  internal node.

The returned vector must be named using the stable internal-node
identifiers supplied in `tree$node.label`.

This interface allows different ancestral-state reconstruction methods
to be connected to the same patching framework through method-specific
adapters.

## Development installation

`phyloPatch` is currently under development and has not yet been released
as a stable public version.

From a local source checkout, the development version can be loaded with:

```r
devtools::load_all(".")
```

## Minimal example

The following example uses a simple toy ancestral-state reconstruction
backend to demonstrate the `phyloPatch` interface.

The toy backend is used only for illustrating package mechanics; it is
not intended as a biological ancestral-state reconstruction method.

```{r example}
library(ape)

tree <- read.tree(
  text = paste0(
    "((((A:1,B:1)J:1,C:1)I:1,D:1)H:1,",
    "(E:1,F:1)K:1)ROOT;"
  )
)

states <- c(
  A = 1,
  B = 2,
  C = 3,
  D = 4,
  E = 5,
  F = 6
)

I_node <- Ntip(tree) +
  match(
    "I",
    tree$node.label
  )

J_node <- Ntip(tree) +
  match(
    "J",
    tree$node.label
  )

shifts <- data.frame(
  parent = I_node,
  child = J_node
)

mean_backend <- function(
    tree,
    states
) {

  stopifnot(
    identical(
      names(states),
      tree$tip.label
    )
  )

  stats::setNames(
    rep(
      mean(states),
      tree$Nnode
    ),
    tree$node.label
  )
}

fit <- patch_asr(
  tree = tree,
  states = states,
  shifts = shifts,
  asr_fun = mean_backend
)

fit
```

The final ancestral-state estimates can be inspected with:

```{r example-results}
fit$ancestral_states
```

The result records both the numerical estimate and its provenance.

For example, component-level ancestral-state estimates are labelled:

```text
source = component_asr
```

whereas reconstructed boundary nodes are labelled:

```text
source = boundary_reconstruction
```

The corresponding component is recorded separately, for example:

```text
mother
patch_1
boundary
```

## Shift-edge specification

Shift edges are supplied as a data frame containing two columns:

```r
data.frame(
  parent = ...,
  child = ...
)
```

Both values refer to the original `ape` node numbers in the exact input
tree supplied to `patch_asr()`.

For example:

```r
shifts <- data.frame(
  parent = 9,
  child = 10
)
```

means that the directed edge:

```text
9 -> 10
```

is treated as the designated branch-localized shift.

Edge-row numbers from `tree$edge` should not be used as node identifiers.

## Output

`patch_asr()` returns an object of class:

```text
patch_asr
```

The main result table is:

```r
fit$ancestral_states
```

and contains:

- stable node identifier;
- original `ape` node number;
- original node label, when available;
- reconstructed ancestral-state estimate;
- computational source of the estimate; and
- component provenance.

Additional information is available in:

```r
fit$patch_plan
fit$boundary_states
fit$component_states
fit$node_map
fit$diagnostics
```

## Methodological boundaries

`phyloPatch` deliberately separates the patching algorithm from the
ancestral-state reconstruction method itself.

The core package therefore handles:

- shift specification;
- stable node identity;
- phylogenetic partitioning;
- component management;
- boundary reconstruction; and
- result merging.

Specific ancestral-state reconstruction methods are connected through
the standardized `asr_fun(tree, states)` interface.

This separation makes it possible to evaluate the patching framework
with different reconstruction methods without changing the core
partitioning algorithm.

## Development status

The package is currently under active development.

The current implementation has automated tests covering:

- input validation;
- stable node identity;
- patch-plan construction;
- internal and terminal patches;
- multiple compatible shifts;
- tree partitioning;
- component-wise reconstruction;
- boundary reconstruction;
- coupled boundary systems;
- result merging; and
- the public `patch_asr()` interface.
